\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
update manuscript

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician and the patient is the risk of an event given their
covariate profile. However, methods relying on time matching or risk-set
sampling (including Cox regression) eliminate the baseline hazard from
the likelihood expression or the estimating function. The baseline
hazard then needs to be estimated separately using a non-parametric
approach. This leads to step-wise estimates of the cumulative incidence
that are difficult to interpret. Using case-base sampling, Hanley \&
Miettinen (2009) explained how the parametric hazard functions can be
estimated using logistic regression. Their approach naturally leads to
estimates of the cumulative incidence that are smooth-in-time.\\
In this paper, we present the \CRANpkg{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis of
time-to-event data. We illustrate all these features through three
different case studies.\\
* SRB and MT contributed equally to this work.
}

@@

In this article, we present the \CRANpkg{casebase} package
\citep{casebase-package} implemented in \textsf{R} \citep{r-core}, for
parametric survival analysis that combines the ideas of Hanley \&
Miettinen \citeyearpar{hanley2009fitting} into a simple interface. The
purpose of the \CRANpkg{casebase} package is to provide practitioners
with an easy-to-use software tool to compute a patient's risk (or
cumulative incidence) of an event, conditional on a particular patient's
covariate profile. Our package retains the flexibility of case-base
sampling and the familiar interface of the \code{glm} function. It also
provides tools for variable selection, competing-risk analysis, and
dealing with time-dependent exposures. In addition, we provide functions
for exploratory data analysis and visualizing the estimated quantities.

In what follows, we first recall some theoretical details on case-base
sampling and its use for estimating parametric hazard functions. We then
give a short review of existing \textsf{R} packages that implement
comparable features as \CRANpkg{casebase}. Next, we provide some details
about the implementation of case-base sampling in our package, and we
give a brief survey of its main functions. This is followed by three
case studies that illustrate the flexibility and capabilities of
\CRANpkg{casebase}. We show how the same framework can be used for
non-linear functions of time and non-proportional hazards, competing
risks and variable selection via penalized regression. Finally, we end
the article with a discussion of the results and of future directions.

@@

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \emph{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
non-informative censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\]

@@

In this model, the instantaneous failure rate is independent of
\(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

Survival analysis is an important branch of applied statistics and
epidemiology. Accordingly, there is already a vast ecosystem of
\textsf{R} packages implementing different methodologies. In this
section, we describe how the functionalities of \CRANpkg{casebase}
compare to these packages.

At the time of writing, a cursory examination of CRAN's \ctv{Survival}
Task View reveals that there are over 250 packages related to survival
analysis \citep{survTaskView}. For the purposes of this article, we
restricted our review to packages that implement at least one of the
following features: parametric modeling, non-proportional hazard models,
competing risk analysis, penalized estimation, and CIF estimation. By
searching for appropriate keywords in the \code{DESCRIPTION} file of
these packages, we found 60 relevant packages. These 60 packages were
then manually examined to determine which ones are comparable to
\CRANpkg{casebase}. In particular, we excluded packages that were
focused on a different set of problems, such as frailty and multi-state
models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs},
along with some of the functionalities they offer.

Parametric survival models are implemented in a handful of packages,
each differing in the underlying data distribution: \CRANpkg{CFC}
\citeyearpar{mahani2015bayesian}, \CRANpkg{flexsurv}
\citeyearpar{flexsurv}, \CRANpkg{SmoothHazard}
\citeyearpar{smoothHazard}, \CRANpkg{rsptm2} \citeyearpar{clements_liu},
\CRANpkg{mets} \citeyearpar{scheike2014estimating}, and
\CRANpkg{survival} \citeyearpar{survival-package}. For example,
\CRANpkg{SmoothHazard} is limited to Weibull distributions
\citeyearpar{smoothHazard}, whereas both \CRANpkg{flexsurv} and
\CRANpkg{survival} allow users to supply any distribution of their
choice. \CRANpkg{flexsurv}, \CRANpkg{smoothhazard}, \CRANpkg{mets} and
\CRANpkg{rstpm2} can model the effect of time using splines, which
allows flexible modeling of the hazard function. As discussed above,
\CRANpkg{casebase} can model any parametric family whose log-hazard can
be expressed as a linear combination of covariates (including time).
Therefore, our package is more general in that it allows the user to
model any linear or non-linear transformation of time including splines
and higher order polynomials. Also, by including interaction terms
between covariates and time, it also allows users to fit
(non-proportional) time-varying coefficient models. However, unlike
\CRANpkg{flexsurv}, we do not explicitly model any shape parameter.

Several packages implement penalized estimation for the Cox model:
\CRANpkg{glmnet} \citeyearpar{regpathcox}, \CRANpkg{glmpath}
\citeyearpar{park_hastie}, \CRANpkg{penalized} \citeyearpar{l1penal},
\CRANpkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\CRANpkg{rstpm2} \citeyearpar{clements_liu}, while \CRANpkg{survival}
\citeyearpar{survival-package} has an implementation of ridge-penalized
estimation. On the other hand, our package \CRANpkg{casebase} provides
penalized estimation of the hazard function. To our knowledge,
\CRANpkg{casebase} and \CRANpkg{rsptm2} are the only packages to offer
this functionality.

Next, several \textsf{R} packages implement methodologies for competing
risk analysis; for a different perspective on this topic, see Mahani \&
Sharabiani \citeyearpar{mahani2015bayesian}. The package
\CRANpkg{cmprsk} provides methods for cause-specific subdistribution
hazards, such as in the Fine-Gray model
\citeyearpar{fine1999proportional}. On the other hand, the package
\CRANpkg{CFC} estimates cause-specific CIFs from unadjusted,
non-parametric survival functions. Our package \CRANpkg{casebase} also
provides functionalities for competing risk analysis by estimating
parametrically the cause-specific hazards. From these quantities, we can
then estimate the cause-specific CIFs.

Finally, several packages include functions to estimate the CIF. The
corresponding methods generally fall into two categories: transformation
of the estimated hazard function, and semi-parametric estimation of the
baseline hazard. The first category broadly corresponds to parametric
survival models, where the full hazard is explicitly modeled. Using this
estimate, the survival function and the CIF can be obtained using their
functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
below). Packages providing this functionality include \CRANpkg{CFC},
\CRANpkg{flexsurv}, \CRANpkg{mets}, and \CRANpkg{survival}. Our package
\CRANpkg{casebase} also follows this approach for both single-event and
competing-risk analyses. The second category outlined above broadly
corresponds to semi-parametric models. These models do not model the
full hazard function, and therefore the baseline hazard needs to be
estimated separately in order to estimate the survival function. This is
achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
parametric estimators (e.g.~spline functions). Packages that implement
this approach include \CRANpkg{riskRegression}, \CRANpkg{rstpm2}, and
\CRANpkg{survival}. As mentioned in the introduction, a key
distinguishing factor between these two approaches is that the first
category leads to smooth estimates of the cumulative incidence function,
whereas the second category often produces estimates in the form of
stepwise functions. Providing smooth estimates of the CIF was one of the
main motivations for introducing case-base sampling in survival
analysis.

@@

\caption{Comparison of various \textsf{R} packages for survival analysis based on several defining features. \textbf{Competing Risks}: whether or not an implementation for competing risks is present. \textbf{Allows Non PH}: permits models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: permits a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Absolute Risk}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}

@@

To illustrate these functions, we will use data from the European
Randomized Study of Prostate Cancer Screening (ERSPC)
\citep{schroder2009screening} which was extracted using the approach
described in Liu \emph{et al.} \citeyearpar{liu2014recovering}. This
dataset is available through the \CRANpkg{casebase} package. It contains
the individual observations for 159,893 men from seven European
countries, who were between the ages of 55 and 69 years when recruited
for the trial.

@@

\begin{itemize}
\tightlist
\item
  \code{glm}: This is the familiar logistic regression.
\item
  \code{glmnet}: This option allows for variable selection using the
  elastic-net \citep{zou2005regularization} penalty (cf.~Case Study 3).
  This functionality is provided through the \CRANpkg{glmnet} package
  \citep{friedman2010jss}.
\item
  \code{gam}: This option provides support for \emph{Generalized
  Additive Models} via the \CRANpkg{mgcv} package
  \citep{hastie1987generalized}.
\item
  \code{gbm}: This option provides support for \emph{Gradient Boosted
  Trees} via the \CRANpkg{gbm} package. This feature is still
  experimental.
\end{itemize}

@@

Previous re-analyses of these data suggested that the overall screening
attributed reduction in death due to prostate cancer of 20\% was an
underestimate \citep{hanley2010mortality}. The estimated 20\% (from a
proportional hazards model) did not account for the delay between
screening and the time the effect is expected to be observed. As a
result, the null effects in years 1--7 masked the substantial reductions
that began to appear from year 8 onward. This motivates the use of a
time-dependent hazard ratio which can easily be fit with the
\CRANpkg{casebase} package by including an interaction term with time in
the model:

@@

In this case study, we show how case-base sampling can be used in the
context of a competing-risk analysis. For illustrative purposes, we use
the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is now available as part of the
\CRANpkg{casebase} package.

@@

For the third case study, we show how \CRANpkg{casebase} can also be
used for variable selection through regularized estimation of the hazard
function as given by \eqref{eq:penest}. To illustrate this
functionality, we use the dataset from the Study to Understand Prognoses
Preferences Outcomes and Risks of Treatment (SUPPORT)
\citep{knaus1995support}. The SUPPORT dataset tracks death in five
American hospitals within individuals who are considered seriously
ill.~The original data is available online from the Department of
Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
and imputed data consists of 9104 observations and 30 variables, and it
is available as part of the \CRANpkg{casebase} package. In the
comparisons below, all covariates except \code{sps} and \code{aps} were
used. These two variables correspond to scores for predicting the
outcome that were developed as part of the original study. For more
information about this dataset, the reader is encouraged to look at the
documentation in our package.

For our penalized case-base model, we opt for the natural log of time
which corresponds to a Weibull distribution. For fitting the penalized
hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
to the \code{fitSmoothHazard} function. We supply both a matrix \code{y}
containing the time and event variables, and a matrix \code{x}
containing all other covariates. We apply the lasso penalty by setting
\code{alpha = 1} and assign a \code{penalty.factor} (\(w_j\);
cf.~Equation \ref{eq:penest}) of 0 to the time variable to ensure it is
in the selected model. We compare our approach to both Cox regression,
and lasso penalized Cox regression (fitted via the \CRANpkg{glmnet}
package).

To compare the performance of our models, we split the data into 95\%
training and 5\% test sets. To assess both discrimination and
calibration, we use a time-dependent version of the classical Brier
score that is adjusted for censoring \citep{graf1999ass}. The Brier
score can be used with both parametric and semi-parametric models. We
use the \CRANpkg{riskRegression} package to compute these scores for all
models.

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \CRANpkg{glmnet} package (Pen. Cox), and our approach using penalized case-base sampling (Pen. CB). Only the covariates that were selected by both penalized approaches are shown. The shrinkage of the coefficient estimates for Pen. Cox and Pen. CB occurs due to the $\ell_1$ penalty.}\label{fig:coefplots}
\end{figure}
\end{Schunk}

@@

In this article, we presented the \textsf{R} package \CRANpkg{casebase},
which provides functions for fitting smooth parametric hazards and
estimating CIFs using case-base sampling. Our package also provides
several functions to produce graphical summaries of the data and the
results. We outlined the theoretical underpinnings of the approach, we
provided details about our implementation, and we illustrated the merits
of the approach and the package through three case studies.

@@

Even though we did not illustrate it in this article, case-base sampling
can also be combined with the framework of \emph{generalized additive
models}. This functionality has already been implemented in the package.
Similarly, case-base sampling can be combined with quasi-likelihood
estimation to fit survival models that can account for the presence of
over-dispersion. All of these examples illustrate how the case-base
sampling framework in general, and the package \CRANpkg{casebase} in
particular, allows the user to fit a broad and flexible family of
survival functions.

@@

In the case studies above, we compared the performance of case-base
sampling with that of Cox regression and Fine-Gray models. In terms of
function interface, \CRANpkg{casebase} uses a formula interface that is
closer to that of \code{glm}, in that the event variable is the only
variable appearing on the left-hand side of the formula. By contrast,
both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
that capture both the event type and time. Both approaches to modeling
yield user-friendly code. However, in terms of output, both approaches
differ significantly. Case-base sampling produces smooth hazards and
smooth cumulative incidence curves, whereas Cox regression and Fine-Gray
models produce stepwise CIFs and never explicitly model the hazard
function. Qualitatively, we showed that by using splines in the linear
predictor, all three models yielded similar curves. However, the smooth
nature of the output of \CRANpkg{casebase} provides a more intuitive
interpretation for consumers of these predictions. In Table
\ref{tab:compCBvsCox}, we provide a side-by-side comparison between the
Cox model and case-base sampling.

\begin{table}
\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
\centering
\begin{tabular}[t]{llp{5cm}}
\toprule
Feature & Cox model & Case-base sampling\\
\toprule
Model type & Semi-parametric & Fully parametric\\\midrule
Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule
Cumulative incidence & Step function & Smooth-in-time curve\\\midrule
Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule
Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\
 & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule
\addlinespace
Competing risks & Difficult & Cause-specific CIFs\\
% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
\bottomrule
\end{tabular}
\end{table}

Our choice of modeling the log-hazard as a linear function of covariates
allows us to develop a simple computational scheme for estimation.
However, as a downside, it does not allow us to model location and scale
parameters separately like the package \CRANpkg{flexsurv}. For example,
if we look at the Weibull distribution as parametrised in
\code{stats::pweibull}, the log-hazard function is given by
\[ \log \lambda(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
where \(\alpha,\beta\) are shape and scale parameters, respectively.
Unlike \CRANpkg{casebase}, the approach taken by \CRANpkg{flexsurv} also
allows the user to model the scale parameter as a function of
covariates. Of course, this added flexibility comes at the cost of
interpretability: by modeling the log-hazard directly, the parameter
estimates from \CRANpkg{casebase} can be interpreted as estimates of
log-hazard ratios. To improve the flexibility of \CRANpkg{casebase} at
capturing the scale of a parametric family, we could replace the
logistic regression with its quasi-likelihood counterpart and therefore
model over- and under-dispersion with respect to the logistic
likelihood. We defer the study of the properties and performance of such
a model to a future article.

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
would like to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools could then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g.~\CRANpkg{SmoothHazard},
\CRANpkg{survival} and \CRANpkg{rstpm2}) provide functions to account
for it. Again, we hope to include any resulting methodology as part of
this package.

>>>>

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician is their patient's risk of an event. However, methods
relying on time matching or risk-set sampling (including Cox regression)
eliminate the baseline hazard from the estimating function. The baseline
hazard then needs to be estimated separately using a non-parametric
approach. This leads to step-wise estimates of the cumulative incidence
that are difficult to interpret. Using case-base sampling, Hanley \&
Miettinen (2009) explained how parametric hazard functions can be
estimated using logistic regression. Their approach naturally leads to
estimates of the cumulative incidence that are smooth-in-time.\\
In this paper, we present the \CRANpkg{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis. We
illustrate all these features through three different case studies.\\
* SRB and MT contributed equally to this work.
}

@@

In this article, we present the \CRANpkg{casebase} package
\citep{casebase-package} implemented in R, for parametric survival
analysis that combines the ideas of Hanley \& Miettinen
\citeyearpar{hanley2009fitting} into a simple interface. The purpose of
the \pkg{casebase} package is to provide practitioners with an
easy-to-use software tool to compute a patient's risk (or cumulative
incidence) of an event, conditional on a particular patient's covariate
profile. Our package retains the flexibility of case-base sampling and
the familiar interface of the \code{glm} function. It also provides
tools for variable selection, competing-risk analysis, and dealing with
time-dependent exposures. In addition, we provide functions for
exploratory data analysis and visualizing the estimated quantities.

In what follows, we first recall some theoretical details on case-base
sampling and its use for estimating parametric hazard functions. We then
give a short review of existing R packages that implement comparable
features as \pkg{casebase}. Next, we provide some details about the
implementation of case-base sampling in our package, and we give a brief
survey of its main functions. This is followed by three case studies
that illustrate the flexibility and capabilities of \pkg{casebase}. We
show how the same framework can be used for non-linear functions of time
and non-proportional hazards, competing risks and variable selection via
penalized regression. Finally, we end the article with a discussion of
the results and of future directions.

@@

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
non-informative censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\]

@@

In this model, the instantaneous failure rate is independent of
\(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

Survival analysis is an important branch of applied statistics and
epidemiology. Accordingly, there is already a vast ecosystem of R
packages implementing different methodologies. In this section, we
describe how the functionalities of \pkg{casebase} compare to these
packages.

At the time of writing, a cursory examination of CRAN's \ctv{Survival}
Task View reveals that there are over 250 packages related to survival
analysis \citep{survTaskView}. For the purposes of this article, we
restricted our review to packages that implement at least one of the
following features: parametric modeling, non-proportional hazard models,
competing risk analysis, penalized estimation, and CIF estimation. By
searching for appropriate keywords in the \code{DESCRIPTION} file of
these packages, we found 60 relevant packages. These 60 packages were
then manually examined to determine which ones are comparable to
\pkg{casebase}. In particular, we excluded packages that were focused on
a different set of problems, such as frailty and multi-state models. The
remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with
some of the functionalities they offer.

Parametric survival models are implemented in a handful of packages,
each differing in the underlying data distribution: \CRANpkg{CFC}
\citeyearpar{mahani2015bayesian}, \CRANpkg{flexsurv}
\citeyearpar{flexsurv}, \CRANpkg{SmoothHazard}
\citeyearpar{smoothHazard}, \CRANpkg{rstpm2} \citeyearpar{clements_liu},
\CRANpkg{mets} \citeyearpar{scheike2014estimating}, and
\CRANpkg{survival} \citeyearpar{survival-package}. For example,
\pkg{SmoothHazard} is limited to Weibull distributions
\citeyearpar{smoothHazard}, whereas both \pkg{flexsurv} and
\pkg{survival} allow users to supply any distribution of their choice.
\pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can
model the effect of time using splines, which allows flexible modeling
of the hazard function. As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear
combination of covariates (including time). Therefore, our package is
more general in that it allows the user to model any linear or
non-linear transformation of time including splines and higher order
polynomials. Also, by including interaction terms between covariates and
time, it also allows users to fit (non-proportional) time-varying
coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly
model any shape parameter.

Several packages implement penalized estimation for the Cox model:
\CRANpkg{glmnet} \citeyearpar{regpathcox}, \CRANpkg{glmpath}
\citeyearpar{park_hastie}, \CRANpkg{penalized} \citeyearpar{l1penal},
\CRANpkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{rstpm2} \citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation of ridge-penalized
estimation. On the other hand, our package \pkg{casebase} provides
penalized estimation of the hazard function. To our knowledge,
\pkg{casebase} and \pkg{rstpm2} are the only packages to offer this
functionality.

Next, several R packages implement methodologies for competing risk
analysis; for a different perspective on this topic, see Mahani \&
Sharabiani \citeyearpar{mahani2015bayesian}. The package
\CRANpkg{cmprsk} provides methods for cause-specific subdistribution
hazards, such as in the Fine-Gray model
\citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric
survival functions. Our package \pkg{casebase} also provides
functionalities for competing risk analysis by estimating parametrically
the cause-specific hazards. From these quantities, we can then estimate
the cause-specific CIFs.

Finally, several packages include functions to estimate the CIF. The
corresponding methods generally fall into two categories: transformation
of the estimated hazard function, and semi-parametric estimation of the
baseline hazard. The first category broadly corresponds to parametric
survival models, where the full hazard is explicitly modeled. Using this
estimate, the survival function and the CIF can be obtained using their
functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
below). Packages providing this functionality include \pkg{CFC},
\pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
\pkg{casebase} also follows this approach for both single-event and
competing-risk analyses. The second category outlined above broadly
corresponds to semi-parametric models. These models do not model the
full hazard function, and therefore the baseline hazard needs to be
estimated separately in order to estimate the survival function. This is
achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
parametric estimators (e.g.~spline functions). Packages that implement
this approach include \pkg{riskRegression}, \pkg{rstpm2}, and
\pkg{survival}. As mentioned in the introduction, a key distinguishing
factor between these two approaches is that the first category leads to
smooth estimates of the cumulative incidence function, whereas the
second category often produces estimates in the form of stepwise
functions. Providing smooth estimates of the CIF was one of the main
motivations for introducing case-base sampling in survival analysis.

@@

\caption{Comparison of various R packages for survival analysis based on several defining features. \textbf{Competing Risks}: whether or not an implementation for competing risks is present. \textbf{Allows Non PH}: permits models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: permits a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Absolute Risk}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}

@@

To illustrate these functions, we will use data from the European
Randomized Study of Prostate Cancer Screening (ERSPC)
\citep{schroder2009screening} which was extracted using the approach
described in Liu \emph{et al.} \citeyearpar{liu2014recovering}. This
dataset is available through the \pkg{casebase} package. It contains the
individual observations for 159,893 men from seven European countries,
who were between the ages of 55 and 69 years when recruited for the
trial.

@@

\begin{itemize}
\tightlist
\item
  \code{glm}: This is the familiar logistic regression.
\item
  \code{glmnet}: This option allows for variable selection using the
  elastic-net \citep{zou2005regularization} penalty (cf.~Case Study 3).
  This functionality is provided through the \pkg{glmnet} package
  \citep{friedman2010jss}.
\item
  \code{gam}: This option provides support for \emph{Generalized
  Additive Models} via the \CRANpkg{mgcv} package
  \citep{hastie1987generalized}.
\item
  \code{gbm}: This option provides support for \emph{Gradient Boosted
  Trees} via the \CRANpkg{gbm} package. This feature is still
  experimental.
\end{itemize}

@@

Previous re-analyses of these data suggested that the overall screening
attributed reduction in death due to prostate cancer of 20\% was an
underestimate \citep{hanley2010mortality}. The estimated 20\% (from a
proportional hazards model) did not account for the delay between
screening and the time the effect is expected to be observed. As a
result, the null effects in years 1--7 masked the substantial reductions
that began to appear from year 8 onward. This motivates the use of a
time-dependent hazard ratio which can easily be fit with the
\pkg{casebase} package by including an interaction term with time in the
model:

@@

In this case study, we show how case-base sampling can be used in the
context of a competing-risk analysis. For illustrative purposes, we use
the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is now available as part of the
\pkg{casebase} package.

@@

For the third case study, we show how \pkg{casebase} can also be used
for variable selection through regularized estimation of the hazard
function as given by \eqref{eq:penest}. To illustrate this
functionality, we use the dataset from the Study to Understand Prognoses
Preferences Outcomes and Risks of Treatment (SUPPORT)
\citep{knaus1995support}.\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}}
The SUPPORT dataset tracks death in five American hospitals within
individuals who are considered seriously ill. The cleaned and imputed
data consists of 9104 observations and 30 variables, and it is available
as part of the \pkg{casebase} package. In the comparisons below, all
covariates except \code{sps} and \code{aps} were used. These two
variables correspond to scores for predicting the outcome that were
developed as part of the original study. For more information about this
dataset, the reader is encouraged to look at the documentation in our
package.

For our penalized case-base model, we opt for the natural log of time
which corresponds to a Weibull distribution. For fitting the penalized
hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
to the \code{fitSmoothHazard} function. We supply both a matrix \code{y}
containing the time and event variables, and a matrix \code{x}
containing all other covariates. We apply the lasso penalty by setting
\code{alpha = 1} and assign a \code{penalty.factor} (\(w_j\);
cf.~Equation \ref{eq:penest}) of 0 to the time variable to ensure it is
in the selected model. We compare our approach to both Cox regression,
and lasso penalized Cox regression (fitted via the \pkg{glmnet}
package).

To compare the performance of our models, we split the data into 95\%
training and 5\% test sets. To assess both discrimination and
calibration, we use a time-dependent version of the classical Brier
score that is adjusted for censoring \citep{graf1999ass}. The Brier
score can be used with both parametric and semi-parametric models. We
use the \pkg{riskRegression} package to compute these scores for all
models.

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen. Cox), and our approach using penalized case-base sampling (Pen. CB). Only the covariates that were selected by both penalized approaches are shown. The shrinkage of the coefficient estimates for Pen. Cox and Pen. CB occurs due to the $\ell_1$ penalty.}\label{fig:coefplots}
\end{figure}
\end{Schunk}

@@

In this article, we presented the R package \pkg{casebase}, which
provides functions for fitting smooth parametric hazards and estimating
CIFs using case-base sampling. Our package also provides several
functions to produce graphical summaries of the data and the results. We
outlined the theoretical underpinnings of the approach, we provided
details about our implementation, and we illustrated the merits of the
approach and the package through three case studies.

@@

Even though we did not illustrate it in this article, case-base sampling
can also be combined with the framework of \emph{generalized additive
models}. This functionality has already been implemented in the package.
Similarly, case-base sampling can be combined with quasi-likelihood
estimation to fit survival models that can account for the presence of
over-dispersion. All of these examples illustrate how the case-base
sampling framework in general, and the package \pkg{casebase} in
particular, allows the user to fit a broad and flexible family of
survival functions.

@@

In the case studies above, we compared the performance of case-base
sampling with that of Cox regression and Fine-Gray models. In terms of
function interface, \pkg{casebase} uses a formula interface that is
closer to that of \code{glm}, in that the event variable is the only
variable appearing on the left-hand side of the formula. By contrast,
both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
that capture both the event type and time. Both approaches to modeling
yield user-friendly code. However, in terms of output, both approaches
differ significantly. Case-base sampling produces smooth hazards and
smooth cumulative incidence curves, whereas Cox regression and Fine-Gray
models produce stepwise CIFs and never explicitly model the hazard
function. Qualitatively, we showed that by using splines in the linear
predictor, all three models yielded similar curves. However, the smooth
nature of the output of \pkg{casebase} provides a more intuitive
interpretation for consumers of these predictions. In Table
\ref{tab:compCBvsCox}, we provide a side-by-side comparison between the
Cox model and case-base sampling.

\begin{widetable}
\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
\centering
\begin{tabular}[t]{llp{5cm}}
\toprule
Feature & Cox model & Case-base sampling\\
\toprule
Model type & Semi-parametric & Fully parametric\\\midrule
Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule
Cumulative incidence & Step function & Smooth-in-time curve\\\midrule
Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule
Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\
 & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule
\addlinespace
Competing risks & Difficult & Cause-specific CIFs\\
% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
\bottomrule
\end{tabular}
\end{widetable}

Our choice of modeling the log-hazard as a linear function of covariates
allows us to develop a simple computational scheme for estimation.
However, as a downside, it does not allow us to model location and scale
parameters separately like the package \pkg{flexsurv}. For example, if
we look at the Weibull distribution as parametrised in
\code{stats::pweibull}, the log-hazard function is given by
\[ \log \lambda(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
where \(\alpha,\beta\) are shape and scale parameters, respectively.
Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} also allows
the user to model the scale parameter as a function of covariates. Of
course, this added flexibility comes at the cost of interpretability: by
modeling the log-hazard directly, the parameter estimates from
\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
improve the flexibility of \pkg{casebase} at capturing the scale of a
parametric family, we could replace the logistic regression with its
quasi-likelihood counterpart and therefore model over- and
under-dispersion with respect to the logistic likelihood. We defer the
study of the properties and performance of such a model to a future
article.

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
would like to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools could then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g. \pkg{SmoothHazard},
\pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
Again, we hope to include any resulting methodology as part of this
package.
